An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-02-14 Paper/BoundEffRed.tex
colors

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds. {\color{red}Videos of these real-time SSTs are available online at \url{https://github.com/AdMeynard/BoundaryEffectsReduction}.}}
\label{fig:ex.intro}
\end{figure}

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
%
{\color{red}
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}.
}
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

>>>>

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds. Videos of these real-time SSTs are available online at \url{https://github.com/AdMeynard/BoundaryEffectsReduction}.}
\label{fig:ex.intro}
\end{figure}

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
%
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}.
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

### 2021-02-14 Paper/results.tex
colors

{\color{red}
\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.
}

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
%
{\color{red}
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.
}

@@

{\color{red}
\subsubsection{Two-Component Respiratory Signal}
Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also a cardiac component known as the cardiogenic artifact~\cite{Smith94recognition}. The top of Fig.~\ref{fig:resp.2} shows an excerpt of the measured thoracic impedance-based respiratory signal, recorded during a bronchoscopic sedation procedure, along with the {\sf SigExt} extension. Below, the simultaneously recorded ECG signal is depicted. The magenta dotted lines emphasize the coincidences between the cardiogenic artifact of the respiratory signal and the QRS complexes of the ECG signal. Such cardiac information can be harvested and utilized when other first-line heart rate information resource is not available or broken~\cite{Lu19recycling}.

@@

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\includegraphics[width=.48\textwidth]{Resp2compZOOM.eps}
\caption{Ordinary SST (top-left) and boundary-free SST (top-right) of a two-component respiratory signal. Middle and bottom: zoom on the areas delimited by the red dashed rectangles; these images are rotated 90 degrees clockwise. The window length for the SSTs is 20~seconds. The instantaneous frequency of the respiration is indicated by the green arrows, while the instantaneous frequency of the cardiogenic artifact is indicated by the blue arrows.}
\label{fig:resp.2.sst}
\end{figure}
}

>>>>

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
%
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.

@@

\subsubsection{Two-Component Respiratory Signal}
Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also a cardiac component known as the cardiogenic artifact~\cite{Smith94recognition}. The top of Fig.~\ref{fig:resp.2} shows an excerpt of the measured thoracic impedance-based respiratory signal, recorded during a bronchoscopic sedation procedure, along with the {\sf SigExt} extension. Below, the simultaneously recorded ECG signal is depicted. The magenta dotted lines emphasize the coincidences between the cardiogenic artifact of the respiratory signal and the QRS complexes of the ECG signal. Such cardiac information can be harvested and utilized when other first-line heart rate information resource is not available or broken~\cite{Lu19recycling}.

@@

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\includegraphics[width=.48\textwidth]{Resp2compZOOM.eps}
\caption{Ordinary SST (top-left) and boundary-free SST (top-right) of a two-component respiratory signal. Middle and bottom: zoom on the areas delimited by the red dashed rectangles; these images are rotated 90 degrees clockwise. The window length for the SSTs is 20~seconds. The instantaneous frequency of the respiration is indicated by the green arrows, while the instantaneous frequency of the cardiogenic artifact is indicated by the blue arrows.}
\label{fig:resp.2.sst}
\end{figure}
